---
title: "Data Re-Exploration"
author: "David King"
format: pdf
---

# Libraries and Data

```{r Libraries}
library(tidyverse)
```


```{r Load Data}
players <- read.csv("players.csv")
player_valuations <- read.csv("player_valuations.csv")
appearances <- read.csv("appearances.csv")
```


# Data Exploration

## Players

```{r Data-Ex Players}
glimpse(players)

ggplot(players, aes(x = factor(last_season))) +
  geom_bar() # Goes from 2013-2022; highest by a lot in 2022; weird (but understandable) break in 2020
             # Pretty even per year until 2022 - doesn't actually increase that much over time
             # I wonder if 2013 doesn't also encapsulate some earlier years? Need to cross-ref with the actual data source

players %>%
  filter(!is.na(country_of_birth) & country_of_birth != "") %>% 
  count(country_of_birth) %>% 
  arrange(desc(n)) %>% 
  head(15) %>% 
  ggplot(aes(x = n, y = reorder(country_of_birth, n))) +
  geom_col()
# All the pretty obvious western european countries
# Intriguingly, the USSR still counts - I wonder if there's an issue where the older players in this dataset were born during the USSR and the later weren't
# UK is split up into its countries


players %>%
  filter(!is.na(country_of_citizenship) & country_of_citizenship != "") %>% 
  count(country_of_citizenship) %>% 
  arrange(desc(n)) %>% 
  head(15) %>% 
  ggplot(aes(x = n, y = reorder(country_of_citizenship, n))) +
  geom_col()
# Looks pretty similar to me

hist(as.Date(players$date_of_birth), breaks = 25) # Some old heads, little left-skewed

players %>% 
  filter(!is.na(position) & position != "") %>% 
  count(position) %>% 
  ggplot(aes(x = reorder(position, n), y = n)) +
  geom_col()
# Maybe a few more goalkeepers than I would have expected, but I'm not a soccer expert

players %>% 
  filter(!is.na(position) & position != "") %>% 
  filter(!is.na(sub_position) & sub_position != "") %>% 
  count(position, sub_position) %>% 
  ggplot(aes(x = sub_position, y = n)) +
  geom_col() +
  facet_wrap(~position, scales = "free_y") +
  coord_flip()
# Curious how these are calculated - is it primary position? Does it vary by season?

players %>% 
  filter(!is.na(foot) & foot != "") %>% 
  count(foot) %>% 
  ggplot(aes(x = foot, y = n)) +
  geom_col()
# About as expected

hist(players$height_in_cm) # Several players with missing data

hist(players[players$height_in_cm > 0,]$height_in_cm)

players %>% 
  filter(!is.na(market_value_in_gbp)) %>% 
  ggplot(aes(x = market_value_in_gbp)) +
  geom_histogram() # Skewed just how you expect

players %>% 
  filter(!is.na(highest_market_value_in_gbp)) %>% 
  ggplot(aes(x = highest_market_value_in_gbp)) +
  geom_histogram() # Not quite as bad but still bad

players %>% 
  filter(!is.na(market_value_in_gbp)) %>% 
  ggplot(aes(x = log(market_value_in_gbp))) +
  geom_histogram(color = "white") # Almost not skewed

players %>% 
  filter(!is.na(market_value_in_gbp) & !is.na(highest_market_value_in_gbp)) %>% 
  select(market_value_in_gbp, highest_market_value_in_gbp) %>% 
  cor() # r = .8397, which is high but not as high as you might expect

players %>% 
  filter(!is.na(agent_name) & agent_name != "") %>% 
  count(agent_name) %>% 
  arrange(desc(n)) %>% 
  head(15) %>% 
  ggplot(aes(x = n, y = reorder(agent_name, n))) +
  geom_col()

top_agents <- players %>% 
  filter(!is.na(agent_name) & agent_name != "") %>% 
  count(agent_name) %>% 
  arrange(desc(n)) %>% 
  head(15)


# Top earning agents - mean
players %>% 
  filter(agent_name %in% top_agents$agent_name) %>% 
  group_by(agent_name) %>% 
  summarize(n_players = n(),
            avg_mv = mean(market_value_in_gbp, na.rm = T),
            avg_high_mv = mean(highest_market_value_in_gbp, na.rm = T)) %>% 
  ggplot(aes(y = reorder(agent_name, n_players), x = n_players,
             fill = avg_mv)) +
  geom_col() # Gestifute is weirdly high here

players %>% 
  filter(agent_name %in% top_agents$agent_name) %>% 
  group_by(agent_name) %>% 
  summarize(n_players = n(),
            avg_mv = mean(market_value_in_gbp, na.rm = T),
            avg_high_mv = mean(highest_market_value_in_gbp, na.rm = T)) %>% 
  ggplot(aes(y = reorder(agent_name, n_players), x = n_players,
             fill = avg_high_mv)) +
  geom_col() # Gestifute is weirdly high here too


# Top earning agents - median
players %>% 
  filter(agent_name %in% top_agents$agent_name) %>% 
  group_by(agent_name) %>% 
  summarize(n_players = n(),
            med_mv = median(market_value_in_gbp, na.rm = T),
            med_high_mv = median(highest_market_value_in_gbp, na.rm = T)) %>% 
  ggplot(aes(y = reorder(agent_name, n_players), x = n_players,
             fill = med_mv)) +
  geom_col() # Gestifute is still weirdly high here, perhaps even more so

players %>% 
  filter(agent_name %in% top_agents$agent_name) %>% 
  group_by(agent_name) %>% 
  summarize(n_players = n(),
            med_mv = median(market_value_in_gbp, na.rm = T),
            med_high_mv = median(highest_market_value_in_gbp, na.rm = T)) %>% 
  ggplot(aes(y = reorder(agent_name, n_players), x = n_players,
             fill = med_high_mv)) +
  geom_col() # Same here


players %>% 
  filter(!is.na(club_pretty_name) & club_pretty_name != "") %>% 
  count(club_pretty_name) %>% 
  arrange(desc(n)) %>% 
  head(15) %>% 
  ggplot(aes(y = reorder(club_pretty_name, n), x = n)) +
  geom_col()

# Nothing too crazy

# Here's a thought: a club that's paying Player A a lot of money doesn't want to pay Player B as much
# Can we make this a feature in the model? Maybe "total club spend" or something?
# Might also have a non-linear relationship - some clubs are willing to just shell it out, others are not
# Can we find any data on clubs themselves? This could be like a multi-level modeling jaunt
```

TODO: Look at missing data
TODO: correlate counts of birth country & citizenship country (how do they determine citizenship, particularly for dual citizens?)

```{r summary and missing values}
summary(players)

for(i in 1:21){
  testcol <- players[,i]
  testcol <- replace(testcol, testcol == "", NA)
  print(c(colnames(players[i]), sum(is.na(testcol))))
}
```


# Ideas

- Club specifics - are certain clubs willing to spend more than others?
- This could be a Bayesian model - update over time as more information comes in (via the API connection), predict value as it changes over time